SU($n$) ferromagnets near the Heisenberg critical point

### 2020-06-09 collective_sun/sun_perturbation.tex
simplified some text, added citation to XXZ paper, added plots of two-body excitation matrices

\DeclareMathOperator{\diag}{diag}

@@

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control,
thereby enabling the detailed study of these symmetries' consequences
for fundamental questions in physics, as well as their practical use
in technological applications.  SU(2)-symmetric spin interactions, for
example, can play an important role in the development of quantum
sensors that make use of entanglement to surpass classical limits on
measurement precision\cite{he2019engineering}\red{[cite XXZ squeezing
  paper]}.  The exciting prospect of similarly exploiting larger
SU($n$) symmetries to achieve a technological advantage is still an
unexplored avenue of research with enormous potential.

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute matrix
elements of arbitrary spin operators within manifolds of weakly broken
permutational symmetry, thereby enabling the numerical simulation of
dynamics induced by non-perturbative modifications to the SU($n$)
Heisenberg model.  While the latter simulations generally accumulate
uncontrolled truncation errors, we identify criteria under which
numerical results correctly capture the qualitative dynamical behavior
of a system.

In the following section, \ref{sec:bare_sun}, we introduce the SU($n$)
Heisenberg model, finding a convenient way to express its Hamiltonian
in terms of SWAP operations that are agnostic to the value of $n$.  In
Section \ref{sec:breaking_sun}, we consider modifying the Hamiltonian
of the bare Heisenberg model by the addition of spin operators,
corresponding to e.g.~external fields or additional interactions, that
weakly break SU($n$) symmetry.  We sketch out a derivation of the
effective Hamiltonian induced by these terms through second order in
perturbation theory, deferring more detailed calculations to
Appendices \ref{sec:eigenstates} and \ref{sec:operator_product}.
Finally, we discuss a numerical method to simulate dynamics within
manifolds of weak permutational symmetry breaking in Section
\ref{sec:shell_model}, before concluding in Section
\ref{sec:conclusions}.  Wherever possible, we will fall back to the
broadly familiar case of SU(2) in order to facilitate an understanding
of the generalizations to SU($n$).

@@

As a concrete example, this sort of scenario can be used in the SU(2)
case to generate spin squeezed states that harness entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering}\red{[cite XXZ squeezing paper]}.
Starting with a general (gapped) SU(2)-symmetric Hamiltonian of the
form in \eqref{eq:H_SU2} we can consider the addition of Ising
interactions
\begin{align}
  H_{\t{Ising}} = \sum_{i\ne j} J_{ij} s_\z^{(i)} s_\z^{(j)}.
  \label{eq:Ising}
\end{align}
If these interactions are sufficiently weak for perturbative
treatment, as in the case of the XXZ model near the Heisenberg
critical point, then the effective Hamiltonian induced by
$H_{\t{Ising}}$ on initially PS states $\ket\psi\in\M_0$ is the
one-axis twisting (OAT) Hamiltonian\cite{ma2011quantum}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where the collective operator $S_\z\equiv\sum_{j=1}^Ns_\z^{(j)}$, and
the squeezing strength $\chi$ is equal to the mean Ising coupling
strength $J_{ij}$.  If Ising interactions are unavailable, the same
effect can be achieved by applying an inhomogeneous magnetic field,
\begin{align}
  H_{\t{B}} = \sum_i B_i s_\z^{(i)},
  \label{eq:magnetic}
\end{align}
which gives rise to an OAT Hamiltonian at second order in perturbation
theory, with a squeezing strength $\chi$ that is related to the
variance of the magnetic field strength $B_i$.

Similarly to single- and two-body operators such as $H_{\t{B}}$ and
$H_{\t{Ising}}$ in \eqref{eq:magnetic} and \eqref{eq:Ising}, it will
be necessary for our purposes to additionally consider the
introduction of $M$-body operators with $M>2$ \red{[todo: explain why
  better]}. Generalizing the above examples to the case of SU($n$)
systems, we therefore consider a general $M$-body operator of the form
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:multi_body_op}
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator, e.g.~$s_\z\otimes s_\z$ in the case of Ising interactions
with $M=2$; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the
individual spins $k_i\in\ZZ_N$ that the operator
$X_k\equiv X_{k_1k_2\cdots k_M}$ acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M : \t{all entries $k_i$ of $k$ are distinct } },
  \label{eq:off_diags}
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$, which is necessary
to identify for a consistent definition of $X_k$ as an $M$-body
operator.  In this notation, we can write the magnetic field and Ising
Hamiltonians in \eqref{eq:magnetic} and \eqref{eq:Ising} as
$H_{\t{B}}=\O\p{\m B,s_\z}$ and
$H_{\t{Ising}}=\O\p{\m J,s_\z\otimes s_\z}$.

If the addition $\O\p{\m w,X}$ to the SU($n$)-symmetric Hamiltonian
$H_0$ in \eqref{eq:H_0} is sufficiently small, namely with operator
norm $\norm{\O\p{\m w,X}}$ less than half the spectral gap
$\Delta_{\t{gap}}$ of $H_0$, $\norm{\O\p{\m w,X}}<\Delta_{\t{gap}}/2$,
then we can treat the effect of $\O\p{\m w,X}$ on the ground-state PS
manifold $\M_0$ perturbatively.  The effective Hamiltonians
$H_{\t{eff}}^{(1)}$ and $H_{\t{eff}}^{(2)}$ induced by $\O\p{\m w,X}$
on $\M_0$ at first and second order in perturbation theory
are\cite{bravyi2011schrieffer}
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 \O\p{\m w,X} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O\p{\m w,X} \P_\Delta \O\p{\m w,X} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects
$\O\p{\m w,X}$ onto the PS manifold $\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\mean w$ is the average of all $w_k$; and $\col X$ is a
collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} X_k,
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.  In the case of a
magnetic field $s_\z$ or Ising interactions $s_\z\otimes s_\z$, for
example,
\begin{align}
  \col{s_\z} = \sum_i s_\z^{(i)} = S_\z,
  &&
  \col{s_\z\otimes s_\z}
  = \sum_{i\ne j} s_\z^{(i)} s_\z^{(j)}
  = S_\z^2 - \f14 N.
\end{align}
The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff} takes more work to simplify due to the presence of a
projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with
excitation energy $\Delta$.  This projector essentially picks off the
part of $\O\p{\m w,X}$ that is strictly off-diagonal with respect to
the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.  We
therefore need to decompose $\O\p{\m w,X}$ into components that
generate states of definite excitation energy when acting on PS states
$\ket\psi\in\M_0$.  The SU($n$) symmetry of $H_0$ enables such a
decomposition of the form
\begin{align}
  H_0 \O\p{\m w,X} \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O\p{\m w_\Delta,X} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{ij}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $\O\p{\m w,X}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem, as we
discuss in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $\O\p{\m w,X}$ into terms
$\O\p{\m w_\Delta,X}$ that generate states of definite excitation
energy $\Delta$, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \P_0 \O\p{\m w_\Delta,X}^2 \P_0.
\end{align}
Simplifying the product $\P_0 \O\p{\m w_\Delta,X}^2 \P_0$ generally
yields $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator, then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\m w_\Delta\c\m w_\Delta}{N\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}}.
  \label{eq:H_eff_2_1}
\end{align}
In the SU(2) case, this result implies that an inhomogeneous magnetic
field ($X=s_\z$) yields the OAT Hamiltonian $H_{\t{OAT}}$ in
\eqref{eq:OAT} at second order perturbation theory.

@@

\red{[note: this section is not yet finished or simplified]}

In order to find the effective Hamiltonian $H_{\t{eff}}^{(2)}$ induced
by the multi-body operator $\hat\O$ on the PS manifold $\M_0$ at
second order in perturbation theory, we constructed states with
definite excitation energy $\Delta$ with respect to $H_0$.  These
states are generated by the action of an $M$-body operator
$X_{\m w_\Delta}$ on a PS state $\ket\psi\in\M_0$.  We denote the
manifold of all states that can be generated by the action of some
$M$-body operator, but not any $r$-body operator with $r<M$, by
$\M_M$; we will refer to $\M_M$ as the $M$-body excitation manifold.
We then denote the a direct sum of all $\M_r$ for $r\le M$ by
$\col\M_M$, i.e.~$\col\M_M\equiv\bigoplus_{r\le M}\M_r$.  The manifold
$\col\M_M$ consists of all states accessible by the action of $r$-body
operators with $r\le M$ on PS states in $\M_0$.  These definitions are
sketched out in Figure \ref{fig:manifold_sketch}.

@@

Both perturbation theory and beyond-perturbative simulations of the
dynamics induced by an operator $\O\p{\m w,X}$ require us to decompose
$\O\p{\m w,X}$ into a sum of terms that generate eigenstates of the
SU($n$)-symmetric interaction Hamiltonian $H_0$ in \eqref{eq:H_0} when
acting on permutationally symmetric (PS) states $\ket\psi\in\M_0$.  To
perform this decomposition, in this section we simplify products of
the form $H_0\O\p{\m w,X}\ket\psi$ with $\ket\psi\in\M_0$.  For
reference, the $M$-body operator $\O\p{\m w,X}$ is defined by
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the individual
spins $k_i\in\ZZ_N$ that the operator $X_k\equiv X_{k_1k_2\cdots k_M}$
acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M: \t{all entries $k_i$ of $k$ are distinct } },
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$.

@@

>>>>

\newcommand{\oper}{\operatorname}

@@

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{gorshkov2010twoorbital, beverland2016realizing,
  cazalilla2014ultracold, taie2012su, hofrichter2016direct,
  cappellini2014direct, scazza2014observation, zhang2014spectroscopic,
  goban2018emergence, perlin2019effective} through the independence of
atomic orbital and interaction parameters on the $n$ nuclear spin
states of individual atoms, with e.g.~$n=10$ in the case of
${}^{87}$Sr.  As a result, experiments can directly probe the SU($n$)
Hubbard model, leading to experimental observations of SU($n$) Hubbard
phases and phase transitions\cite{taie2012su, hofrichter2016direct},
two-orbital SU($n$) magnetism\cite{cappellini2014direct,
  scazza2014observation, zhang2014spectroscopic}, and multi-body
SU($n$)-symmetric interactions\cite{goban2018emergence,
  perlin2019effective}.  The tremendous theoretical significance of
SU($n$) symmetries makes it all the more exciting that they appear in
experimental platforms with exquisite degrees of microscopic control.
This control enables detailed studies of these symmetries'
consequences for fundamental questions in physics, as well as their
practical use in technological applications.  SU(2)-symmetric spin
interactions, for example, can play an important role in the
development of quantum sensors that make use of entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering, perlin2020spin}.
The exciting prospect of similarly exploiting larger SU($n$)
symmetries to achieve a technological advantage is still an unexplored
avenue of research with enormous potential.

\red{[Paragraph on collective spin systems]}

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry (e.g.~due to external fields or additional interaction
terms).  The restriction of our theory to ferromagnetic interactions
can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute matrix
elements of arbitrary spin operators within manifolds of weakly broken
permutational symmetry, thereby enabling the numerical simulation of
dynamics induced by non-perturbative modifications to the SU($n$)
Heisenberg model.  While these simulations generally accumulate
uncontrolled errors from the truncation of Hilbert space, we identify
criteria under which numerical results correctly capture the
qualitative dynamical behavior of a system.

In the following section, \ref{sec:bare_sun}, we introduce the SU($n$)
Heisenberg model, finding a convenient way to express its Hamiltonian
in terms of permutation (or SWAP) operations that are agnostic to the
value of $n$.  In Section \ref{sec:breaking_sun}, we consider
modifying the Hamiltonian of the bare Heisenberg model by the addition
of spin operators, corresponding to e.g.~external fields or additional
interactions, that weakly break SU($n$) symmetry.  We sketch out a
derivation of the effective Hamiltonian induced by these terms through
second order in perturbation theory, deferring more detailed
calculations to appendices.  Finally, we discuss a numerical method to
simulate dynamics within manifolds of weak permutational symmetry
breaking in Section \ref{sec:shell_model}, before concluding in
Section \ref{sec:conclusions}.  Wherever possible, we will fall back
to the broadly familiar case of SU(2) in order to facilitate an
understanding of the generalizations to SU($n$).

@@

As a concrete example, this sort of scenario can be used in the SU(2)
case to generate spin squeezed states that harness entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering, perlin2020spin}.
Starting with a general (gapped) SU(2)-symmetric Hamiltonian of the
form in \eqref{eq:H_SU2} we can consider the addition of Ising
interactions
\begin{align}
  H_{\t{Ising}} = \sum_{i\ne j} J_{ij} s_\z^{(i)} s_\z^{(j)}.
  \label{eq:Ising}
\end{align}
If these interactions are sufficiently weak for perturbative
treatment, as in the case of the XXZ model near the Heisenberg
critical point, then the effective Hamiltonian induced by
$H_{\t{Ising}}$ on initially PS states $\ket\psi\in\M_0$ is the
one-axis twisting (OAT) Hamiltonian\cite{ma2011quantum}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where the collective operator $S_\z\equiv\sum_{j=1}^Ns_\z^{(j)}$, and
the squeezing strength $\chi$ is equal to the mean Ising coupling
strength $J_{ij}$.  If Ising interactions are unavailable, the same
effect can be achieved by applying an inhomogeneous magnetic field,
\begin{align}
  H_{\t{B}} = \sum_i B_i s_\z^{(i)},
  \label{eq:magnetic}
\end{align}
which gives rise to an OAT Hamiltonian at second order in perturbation
theory, with a squeezing strength $\chi$ that is related to the
variance of the magnetic field strength $B_i$.

Similarly to single- and two-body operators such as $H_{\t{B}}$ and
$H_{\t{Ising}}$ in \eqref{eq:magnetic} and \eqref{eq:Ising}, it will
be necessary for our purposes to additionally consider the
introduction of $M$-body operators with $M>2$. Generalizing the above
examples to the case of SU($n$) systems, we therefore consider a
general $M$-body operator of the form
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:multi_body_op}
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator, e.g.~$s_\z\otimes s_\z$ in the case of Ising interactions
with $M=2$; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the
individual spins $k_i\in\ZZ_N$ that the operator
$X_k\equiv X_{k_1k_2\cdots k_M}$ acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M : \t{all entries $k_i$ of $k$ are distinct} },
  \label{eq:off_diags}
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$, which is necessary
to identify for a consistent definition of $X_k$ as an $M$-body
operator.  In this notation, we can write the magnetic field and Ising
Hamiltonians in \eqref{eq:magnetic} and \eqref{eq:Ising} as
$H_{\t{B}}=\O\p{\m B,s_\z}$ and
$H_{\t{Ising}}=\O\p{\m J,s_\z\otimes s_\z}$.

If the addition $\O\p{\m w,X}$ to the SU($n$)-symmetric Hamiltonian
$H_0$ in \eqref{eq:H_0} is sufficiently small, namely with operator
norm $\norm{\O\p{\m w,X}}$ less than half the spectral gap
$\Delta_{\t{gap}}$ of $H_0$, $\norm{\O\p{\m w,X}}<\Delta_{\t{gap}}/2$,
then we can treat the effect of $\O\p{\m w,X}$ on the ground-state PS
manifold $\M_0$ perturbatively.  The effective Hamiltonians
$H_{\t{eff}}^{(1)}$ and $H_{\t{eff}}^{(2)}$ induced by $\O\p{\m w,X}$
on $\M_0$ at first and second order in perturbation theory
are\cite{bravyi2011schrieffer}
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 \O\p{\m w,X} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O\p{\m w,X} \P_\Delta \O\p{\m w,X} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects
$\O\p{\m w,X}$ onto the PS manifold $\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\mean w$ is the average of all $w_k$; and $\col X$ is a
collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} X_k,
  \label{eq:mean_col}
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.  In the case of a
magnetic field $s_\z$ or Ising interactions $s_\z\otimes s_\z$, for
example,
\begin{align}
  \col{s_\z} = \sum_i s_\z^{(i)} = S_\z,
  &&
  \col{s_\z\otimes s_\z}
  = \sum_{i\ne j} s_\z^{(i)} s_\z^{(j)}
  = S_\z^2 - \f14 N.
\end{align}
The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff} takes more work to simplify due to the presence of a
projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with
excitation energy $\Delta$.  This projector essentially picks off the
part of $\O\p{\m w,X}$ that is strictly off-diagonal with respect to
the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.  We
therefore need to decompose $\O\p{\m w,X}$ into components that
generate states of definite excitation energy when acting on PS states
$\ket\psi\in\M_0$.  The SU($n$) symmetry of $H_0$ enables such a
decomposition to take the form
\begin{align}
  H_0 \O\p{\m w,X} \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O\p{\m w_\Delta,X} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{ij}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $\O\p{\m w,X}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem, as we
discuss in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $\O\p{\m w,X}$ into terms
$\O\p{\m w_\Delta,X}$ that generate states of definite excitation
energy $\Delta$, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \P_0 \O\p{\m w_\Delta,X}^2 \P_0.
\end{align}
Simplifying the product $\P_0 \O\p{\m w_\Delta,X}^2 \P_0$ generally
yields $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator, then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\m w_\Delta\c\m w_\Delta}{\Delta} \,
  \mean{X\otimes X}
  = \sum_{\Delta\ne0} \f{\m w_\Delta\c\m w_\Delta}{N\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}},
  \label{eq:H_eff_2_1}
\end{align}
where $\mean{X\otimes X}$ is the average of
$\p{X\otimes X}_{ij}=X_iX_j$ over all $i\ne j$, similarly to the
definition of $\mean{w}$ in Eq.~\eqref{eq:mean_col}.  In the SU(2)
case, this result implies that an inhomogeneous magnetic field
($X=s_\z$) yields the OAT Hamiltonian $H_{\t{OAT}}$ in \eqref{eq:OAT}
at second order perturbation theory.

In order to find the effective Hamiltonian $H_{\t{eff}}^{(2)}$ induced
by the $M$-body operator $\O\p{\m w,X}$ on the PS manifold $\M_0$ at
second order in perturbation theory, we constructed states with
definite excitation energy $\Delta>0$ with respect to $H_0$.  Rather
than treating these excited states as ``intermediate'' or ``virtual''
in the derivation of an effective ground-state Hamiltonian, we can
perform numerical simulations in the joint Hilbert space of both
ground and excited states.  When the perturbation $\O\p{\m w,X}$ is
small, namely with operator norm
$\norm{\O\p{\m w,X}}<\Delta_{\t{gap}}/2$, performing such simulations
with an appropriate choice of excited states is equivalent to
high-order ground-state perturbation theory.  The corresponding order
of perturbation theory is determined by the choice of excited states
in a manner that will be clarified below.  Moreover, the first-class
treatment of excited states enables simulating the dynamics induced by
{\it non-perturbative} additions $\O\p{\m w,X}$ to the base
Hamiltonian $H_0$ in finite systems.

In order to simulate dynamics within excited-state manifolds, we first
need to classify the set of excited states in order to construct a
suitable basis.  In the case of SU(2), a good starting point is to
identify states with definite (non-negative) spin length
$S=\set{N/2,N/2-1,\cdots}$, and spin projection
$m_\z\in\set{S,S-1,\cdots,-S}$ onto a quantization axis.  To
generalize this classification to SU($n$), we denote the manifold of
states with definite fixed spin length $S$ by $\M_{N/2-S}$.  The PS
manifold of Dicke states with $S=N/2$ is then $\M_0$, the manifold of
single-magnon spin-wave states with $S=N/2-1$ is $\M_1$, and more
generally the manifold of $r$-magnon states with $S=N/2-r$ is $\M_r$.
Similarly, rather than the spin projection $m_\z$ we identify states
by labels $m=\p{m_\up,m_\dn}$ in which $m_\mu$ denotes the occupation
number of the single-spin state $\mu$, with $m_\up-m_\dn=m_\z$ and
$m_\up+m_\dn=N$.

The label $m$ suffices to identify the PS (Dicke) manifold
$\M_0=\oper{span}\set{\ket{m}}$, but we need additional labels to
uniquely identify states in the $r$-magnon manifold $\M_r$ with $r>0$.
Single-magnon ($r=1$) spin-wave states of SU(2) spin systems can be
written in the form $\ket{m,\m v}\equiv\O\p{\m v,\sigma_\z}\ket{m}$,
where $\m v=\p{v_1,v_2,\cdots,v_N}$ and $\sigma_\z$ is a Pauli-$z$
operator.  If an SU(2) spin system is translationally invariant, then
its spin-wave eigenstates have $v_j=e^{\i k_{\m v}\c j}/\sqrt{N}$ with
wavenumber $k_{\m v}\ne0$.  Similarly, $r$-magnon states for SU(2)
take the form
$\ket{m,\m v}\equiv\O\p{\m v,\sigma_\z^{\otimes r}}\ket{m}$, where now
$\m v$ is an $r$-index tensor that encodes spatial structure and
correlations in the excited state $\ket{m,\m v}$ (see Figure
\ref{fig:pair_structure}).

\begin{figure}
  \centering
  \includegraphics{pair_structure/pair_L16_A3_s1.pdf}
  \includegraphics{pair_structure/pair_L16_A3_s2.pdf}
  \includegraphics{pair_structure/pair_L16_A3_s3.pdf} \\[1em]
  \includegraphics{pair_structure/pair_L16_A3_s4.pdf}
  \includegraphics{pair_structure/pair_L16_A3_s5.pdf}
  \includegraphics{pair_structure/pair_L16_A3_s6.pdf}
  \includegraphics{pair_structure/pair_L16_A3_s7.pdf}
  \caption{Matrices $\m v_\Delta=\sum_{i,j}v^{(\Delta)}_{ij}\op{i}{j}$
    that generate two-magnon states
    $\ket{\m v_\Delta,m} \equiv \O\p{\m v_\Delta,
      \sigma_\z\otimes\sigma_\z} \ket{m}$ with definite excitation
    energy $\Delta$ in a periodic 1D chain of $N=16$ spins with
    power-law SU(2) couplings $h_{ij}=-1/\oper{dist}\p{i,j}^3$, where
    $\oper{dist}\p{i,j}\equiv\min\p{\abs{i-j},N-\abs{i-j}}$.  Darker
    colors correspond to larger values of $\abs*{v^{(\Delta)}_{ij}}$,
    and a blue (red) color corresponds to a positive (negative) sign
    of $v^{(\Delta)}_{ij}$.  In the case of SU($n$) with $n>2$, the
    ``two-magnon'' excitation energies $\Delta$ and matrices
    $\m v_\Delta$ remain identical, but now correspond to a degenerate
    set of states that generated by different choice of two-body
    operators in place of $\sigma_\z\otimes\sigma_\z$.}
  \label{fig:pair_structure}
\end{figure}

\red{[note: this section is not yet finished]}

@@

\red{[note: text below is old and unsimplified]}

We denote the manifold of all states that can be generated by the
action of some $M$-body operator, but not any $r$-body operator with
$r<M$, by $\M_M$; we will refer to $\M_M$ as the $M$-body excitation
manifold.  We then denote the a direct sum of all $\M_r$ for $r\le M$
by $\col\M_M$, i.e.~$\col\M_M\equiv\bigoplus_{r\le M}\M_r$.  The
manifold $\col\M_M$ consists of all states accessible by the action of
$r$-body operators with $r\le M$ on PS states in $\M_0$.  These
definitions are sketched out in Figure \ref{fig:manifold_sketch}.

@@

Both perturbation theory and beyond-perturbative simulations of the
dynamics induced by an operator $\O\p{\m w,X}$ require us to decompose
$\O\p{\m w,X}$ into a sum of terms that generate eigenstates of the
SU($n$)-symmetric interaction Hamiltonian $H_0$ in \eqref{eq:H_0} when
acting on permutationally symmetric (PS) states $\ket\psi\in\M_0$.  To
perform this decomposition, in this section we simplify products of
the form $H_0\O\p{\m w,X}\ket\psi$ with $\ket\psi\in\M_0$.  For
reference, the $M$-body operator $\O\p{\m w,X}$ is defined by
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the individual
spins $k_i\in\ZZ_N$ that the operator $X_k\equiv X_{k_1k_2\cdots k_M}$
acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M: \t{all entries $k_i$ of $k$ are distinct} },
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
